Optimal management of a stochastically varying population when policy adjustment is costly

### 2014-08-11 manuscripts/supplement.Rmd
additional analyses

---
title: "Fisheries management when quotas are costly to change"
date: "08/12/2014"
output: 
  revealjs_presentation:
    theme: serif
    self_contained: true
    incremental: true
    transition: none
    pandoc_args: ["--css", "components/style.css"]
author: Carl Boettiger, Michael Bode, James Sanchirico, Jacob LaRiviere, Alan Hastings, Paul Armsworth

notes: |
  Time: 30 minutes with questions
  Presenting at: 2014-12-08 15:10:00 PDT

@@

## Our world is changing

<img src="components/images/global-change.jpg" width="500">

<small> Overpeck+ (2011) doi: [10.1126/science.1197869](http://doi.org/10.1126/science.1197869) </small>

## Our policies are not

<img src="components/images/nochange.jpg" width="700">

<small> Alex Wong (Getty Images) </small>

@@

library("methods")
library("knitr")
basename <- "presentation"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)
# PDF-based figures
opts_chunk$set(dev='pdf')
fig.cache <- TRUE

@@

```{r analysis, include=FALSE}
source("components/analysis.R")
```

----

## Natural populations fluctuate...

```{r ex_fish, fig.cap="", dev=c("svg", "png"), dev.args=list(bg="transparent")}
ex <- filter(dt, replicate=="rep_17" & penalty_fn == "L1" & time < 20)
ggplot(ex) + 
  geom_line(aes(time, fishstock), col="darkblue", lwd=1)
```

@@

```{r ex_harvest, fig.cap="", dev=c("svg", "png"), dependson="ex_fish", dev.args=list(bg="transparent")}
ggplot(ex) + 
  geom_line(aes(time, harvest_alt), col="darkgreen", lwd=1) +
  ylab("Optimal quota policy") 
```

## Real policies: not so much

![](components/images/tuna_cc0.png)

<small> NOAA </small>

## Bluefin Tuna Stocks vs Quota (base TAC), ICCAT 1987 - 2007

```{r Figure_1, fig.cap="Quota on Blue Fin Tuna Catch", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
tuna <- read.csv("components/data/tuna.csv")
tuna<-melt(tuna, id="year")
ggplot(tuna, aes(year, value)) + geom_point() +
  facet_wrap(~variable, ncol=1, scale="free_y") + 
  ylab("Bluefin Tuna (tonnes)") + theme_bw()

@@

<small> base TAC for the stock (before any underages or overages are applied or divided by country)</small>

----

> ## As ecologists, we've always focused on accounting for _ecological_ dynamics

## Not so good at _policy_ dynamics

- We account for _ecological_ dynamics but not _political_ ones
- We don't know the equations of politics (does anyone?)
- Instead, we can investigate the potential impact that costly adjustments have on optimal policy

Model setup
------------

- Focus on _stochastic_ dynamics: thus a constant harvest is not optimal.

- Follow the textbook classic approach: __Reed (1979)__ [10.1016/0095-0696(79)90014-7](http://doi.org/10.1016/0095-0696(79)90014-7)

```{r, fig.cap="",  dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent"), fig.height=3}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "L1" & time < 20) %>%
  select(time, fishstock, harvest_alt) %>%
  gather(variable, stock, -time) 
ggplot(ex) + 
  geom_line(aes(time, stock, col=variable), lwd=1) + 
  scale_color_discrete(labels=c(fishstock = "Stock size", harvest_alt = "Harvest policy"), name="")

Fish population dynamics 
--------------------------

### (state equation)

Population grows under a Beverton-Holt stock-recruitment curve,

$$N_{t+1} = Z_t \frac{A (N_t - h_t)}{1 + B (N_t - h_t)}$$

- subject to multiplicative log-normal growth shocks $Z_t$

Optimization
------------

Select harvest policy that maximizes the Net Present Value (NPV) of the stock:

$$\max_{{\bf h}}\mathbf{ E} ( NPV_{0} )=\max_{{\bf h}} \sum_0^\infty \mathbf{E} \left(\frac{\Pi_0(N_t,h_t)}{(1+\delta)^{t+1}} \right)$$

- Profits from harvesting depend on price and costs $\Pi_0(N_t,h_t) = p h_t -  c_0 E_t(N_t, h_t) $
- Harvest quota $h_t$ in year $t$
- Stock size $N_t$, measured before harvest
- discount rate $\delta$

### Costs of policy adjustment

- We replace $\Pi_0$ in the $NPV_0$ equation with...

"Linear costs"
-------------

$$\Pi_{1}(N_t,h_t, h_{t-1}) = \Pi_0 - c_1  |  h_t - h_{t-1} | $$

- All previous terms, minus: 
- The cost to change the harvest policy (quota): proportional to the size of the change

 "Quadratic costs"
-----------------

$$\Pi_{2}(N_t,h_t, h_{t-1}) = \Pi_0 - c_2 (  h_t - h_{t-1})^2 $$

- Small adjustments are very cheap
- Large adjustments are very expensive
- Closest to typical assumptions of quadratic costs on harvest/effort (but here it is the _change_ in harvest/effort)

 "Fixed costs"
------------

$$\Pi_{3}(N_t,h_t, h_{t-1}) = \Pi_0 - c_3 (1-\mathbb{I}(h_t, h_{t-1}))  $$

- A fixed transaction fee for any change to policy, independent of size (or sign) of the change.

<br/>
<br/>
<small>($\mathbb{I(x,y)}$ is indicator function, equals 1 iff $x=y$, zero otherwise.) </small>

@@

### How do we pick coefficients $c_i$ such that only the functional form and not the overall cost differ?

- All forms will create complete resistance to change if the costs are high enough
- But how do we compare across forms that have different units and different coefficients?

```{r Figure_2, fig.cap="",  dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

----

## Problem defined. Time to compute solutions.

## Implementation:

### Stochastic Dynamic Programming

<img src="components/images/mangel_and_clark.jpg", width="300">

----

<img src="components/images/confusion0.png" width="600">

----

<img src="components/images/confusion1.png" width="600">

----

<img src="components/images/confusion_matrix.png" width="600">

----

<img src="components/images/numbers.png" width="600">

----

<img src="components/images/outcomes_numbers.png" width="600">

----

<img src="components/images/time_grid.png" width="600">

----

<img src="components/images/previous_action.png" width="600">

Effect of policy adjustment costs on optimal quotas and stock sizes 
-------------------------------------------------------------------

What would you predict for...

- Quadratic adjustment costs?
- Linear adjustment costs?
- Fixed adjustment costs?

```{r reex, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent"), fig.height=3}
ggplot(ex) + 
  geom_line(aes(time, stock, col=variable), lwd=1) + 
  scale_color_discrete(labels=c(fishstock = "Stock size", harvest_alt = "Harvest policy"), name="") 
```

@@

```{r quadratic, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "L2") %>%
  select(time, harvest, harvest_alt) %>%
  gather(variable, value, -time) 
ggplot(ex) + 
  geom_line(aes(time, value, col=variable), lwd=1) + 
  ylab("harvest quota") +
  scale_color_discrete(labels=c(harvest = "adjusted policy", harvest_alt = "baseline policy"), name="") 
```

@@

```{r fixed, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "fixed") %>%
  select(time, harvest, harvest_alt) %>%
  gather(variable, value, -time) 
ggplot(ex) + 
  geom_line(aes(time, value, col=variable), lwd=1) + 
  ylab("harvest quota") +
  scale_color_discrete(labels=c(harvest = "adjusted policy", harvest_alt = "baseline policy"), name="") 
```

@@

```{r linear, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "L1") %>%
  select(time, harvest, harvest_alt) %>%
  gather(variable, value, -time) 
ggplot(ex) + 
  geom_line(aes(time, value, col=variable), lwd=1) + 
  ylab("harvest quota") +
  scale_color_discrete(labels=c(harvest = "adjusted policy", harvest_alt = "baseline policy"), name="") 
```

@@

```{r Figure_3, dependson="tidy", fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

```{r Figure_4, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ggplot(subset(figure4_df, statistic != "cross.correlation"), 
       aes(penalty_fraction, value, fill=penalty_fn, col=penalty_fn))  +
  stat_summary(fun.y = mean, 
               fun.ymin = function(x) mean(x) - sd(x), 
               fun.ymax = function(x) mean(x) + sd(x), 
               geom = "ribbon", alpha=0.3, colour=NA) +
  stat_summary(fun.y = mean, geom = "line") + 
  coord_cartesian(xlim = c(0, .3)) + 
  facet_wrap(~ timeseries + statistic, scale="free_y") + 
  scale_color_discrete(labels=relabel, name="Penalty function") + 
  scale_fill_discrete(labels=relabel, name="Penalty function") + 
  xlab("Penalty size (as fraction of NPV0)")
```

@@

## Okay, so the policies "look" different. Does it matter?

## Yes

## Managing under the wrong assumptions

$$NPV_i( h_1^* ) = \sum_{t=0}^\infty (\overbrace{ p h^*_{1,t}-c_0E^*_{1t}}^{[1]}-\overbrace{c_1 | h^*_{1,t}- h^*_{1,t-1}|}^{[2]} )  \frac{1}{(1+\delta)^t}$$

$$NPV_i( h_0^* ) = \sum_{t=0}^\infty (\underbrace{ p h^*_{0,t}-c_0E^*_{0t}}_{[3]}-\underbrace{c_1 | h^*_{0,t}- h^*_{0,t-1}|}_{[4]} )  \frac{1}{(1+\delta)^t}$$

<small>

1. Dockside profits when accounting for adjustment cost 
2. (Anticipated) Fees paid for adjusting the policy
3. Theoretical maximum dockside profits 
4. (Unanticipated) fees for those adjustments

</small>

@@

```{r Figure_5, fig.cap="",  dependson="histogram-calc", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

What is the impact of assuming costs are present when they are not?

- _Very little change to your bottom line_ (regardless of cost structure) <br/><br/>

What is the impact of ignoring costs when they are present?

- _Substantially higher costs_ (particularly for quadratic costs).

@@

```{r Figure_6, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

## What about managing when costs are present...

###  ..but you're using the wrong model??

@@

```{r Figure_7, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
who <- c("penalty_fn", "ignore_fraction", "mismatched_fraction", "reduction")
table2 <- arrange(mismatches_df[who], reduction) 
names(table2) = c("penalty.fn", "ignoring", "mismatched", "reduction")
table2_long <- melt(table2, id = c('penalty.fn', 'reduction'))
table2_long$reduction <- as.factor(table2_long$reduction)

@@

## Mismatched costs are intuitive if we remember the general patterns

- Quadratic ($\Pi_2$) smooths
- Linear ($\Pi_1$) flattens
- Fixed ($\Pi_3$) jumps

Predict which is worse: assuming the wrong cost or no assuming no cost at all, when:

- Costs are assumed linear but in reality quadratic?
- True costs are assumed quadratic but in reality fixed?

## Mismatched costs are intuitive

```{r Figure_7again, fig.cap="", dev=c("svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ggplot(table2_long, aes(penalty.fn, value, fill = variable)) + 
  geom_bar(stat="identity", position="dodge") + 
  facet_wrap(~reduction, scales="free_y") +
  scale_x_discrete(labels = c('L1_L2' = substitute(paste(Pi[1],"_", Pi[2])), 
                              'L2_L1' =  substitute(paste(Pi[2],"_", Pi[1])),
                              'L1_fixed' = substitute(paste(Pi[1],"_", Pi[3])), 
                              'fixed_L1' =  substitute(paste(Pi[3],"_", Pi[1])),
                              'fixed_L2' = substitute(paste(Pi[3],"_", Pi[2])), 
                              'L2_fixed' =  substitute(paste(Pi[2],"_", Pi[3])))) + 
  xlab("Penalty function as: assumed_reality") + theme_tufte(base_size=14)
```

## Conclusions

- Policies reflecting the best ecology may ultimately be impractical
- Novel look at introducing costs to changes in policy
- Details matter: some adjustment costs make policy less smooth, some more smooth
- Solutions can satisfy the constraints of adjustment costs while performing nearly optimally
- Ignoring adjustment costs can result in significant penalties
- There are many benefits to joining a NIMBioS working group

## A 'Pretty Darn Good' Working Group

![](components/images/pdg-group.jpg)

## Acknowledgements

<img src="components/images/nsf.jpg" width="100" height="100"><img src="components/images/noaa.jpg" width="100" height="100"><img src="components/images/sloan.jpg" width="100" height="100">  <br/>

<img src="components/images/nimbios.jpg" width="150" height="75"> 
<img src="components/images/csgf.png" width="150" height="75"> <br/>

<!-- Paid the bills and folks that have supported or otherwise
tolerated this and so many other side projects -->

<small>Slides, data, references, code and more at: http://io.carlboettiger.info/pdg_control</small>

>>>>

---
title: Optimal management of a stochastically varying population when policy adjustment is costly
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig (at) gmail.com
    footnote: Corresponding author
  - name: Michael Bode
    affiliation: melbourne
  - name: James Sanchirico
    affiliation: UCD
  - name: Jacob LaRiviere
    affiliation: UTK
  - name: Alan Hastings
    affiliation: UCD
  - name: Paul Armsworth 
    affiliation: UTK

address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research,
      110 Shaffer Rd,
      Santa Cruz, CA 95050, USA 
  - code: UCD
    address: Department of Environmental Science and Policy, University of California, Davis
  - code: melbourne
    address: University of Melbourne, Australia 
  - code: UTK
    address: University of Tennessee, Knoxville

layout: review,12pt
linenumbers: true

bibliography: components/references.bib
csl: components/ecology.csl
documentclass: components/elsarticle

output:
  pdf_document:
    template: components/elsarticle.latex
    fig_caption: true

@@

@@

library("methods")
library("knitr")
basename <- "supplement"
opts_chunk$set(fig.path = paste("components/figure/", basename, "-", sep=""),
               cache.path = paste("components/cache/", basename, "/", sep=""))
opts_chunk$set(cache = 2)
opts_chunk$set(tidy=FALSE, warning=FALSE, message=FALSE, 
               comment = NA, verbose = TRUE, echo=FALSE)
# PDF-based figures
opts_chunk$set(dev='pdf')
fig.cache <- TRUE

@@

```{r analysis, include=FALSE}
source("components/supp_analysis.R")
```

```{r ex_fish, fig.cap="", dev=c("pdf", "svg", "png"), dev.args=list(bg="transparent")}
ex <- filter(dt, replicate=="rep_17" & penalty_fn == "L1" & time < 20)
ggplot(ex) + 
  geom_line(aes(time, fishstock), col="darkblue", lwd=1)
```

@@

```{r ex_harvest, fig.cap="", dev=c("pdf", "svg", "png"), dependson="ex_fish", dev.args=list(bg="transparent")}
ggplot(ex) + 
  geom_line(aes(time, harvest_alt), col="darkgreen", lwd=1) +
  ylab("Optimal quota policy") 
```

```{r, fig.cap="",  dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent"), fig.height=3}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "L1" & time < 20) %>%
  select(time, fishstock, harvest_alt) %>%
  gather(variable, stock, -time) 
ggplot(ex) + 
  geom_line(aes(time, stock, col=variable), lwd=1) + 
  scale_color_discrete(labels=c(fishstock = "Stock size", harvest_alt = "Harvest policy"), name="")

@@

@@

```{r Figure_2, fig.cap="",  dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

```{r reex, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent"), fig.height=3}
ggplot(ex) + 
  geom_line(aes(time, stock, col=variable), lwd=1) + 
  scale_color_discrete(labels=c(fishstock = "Stock size", harvest_alt = "Harvest policy"), name="") 
```

@@

```{r quadratic, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "L2") %>%
  select(time, harvest, harvest_alt) %>%
  gather(variable, value, -time) 
ggplot(ex) + 
  geom_line(aes(time, value, col=variable), lwd=1) + 
  ylab("harvest quota") +
  scale_color_discrete(labels=c(harvest = "adjusted policy", harvest_alt = "baseline policy"), name="") 
```

@@

```{r fixed, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "fixed") %>%
  select(time, harvest, harvest_alt) %>%
  gather(variable, value, -time) 
ggplot(ex) + 
  geom_line(aes(time, value, col=variable), lwd=1) + 
  ylab("harvest quota") +
  scale_color_discrete(labels=c(harvest = "adjusted policy", harvest_alt = "baseline policy"), name="") 
```

@@

```{r linear, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ex <- dt %>% 
  filter(replicate=="rep_17" & penalty_fn == "L1") %>%
  select(time, harvest, harvest_alt) %>%
  gather(variable, value, -time) 
ggplot(ex) + 
  geom_line(aes(time, value, col=variable), lwd=1) + 
  ylab("harvest quota") +
  scale_color_discrete(labels=c(harvest = "adjusted policy", harvest_alt = "baseline policy"), name="") 
```

@@

```{r Figure_3, dependson="tidy", fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

```{r Figure_4, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
ggplot(subset(figure4_df, statistic != "cross.correlation"), 
       aes(penalty_fraction, value, fill=penalty_fn, col=penalty_fn))  +
  stat_summary(fun.y = mean, 
               fun.ymin = function(x) mean(x) - sd(x), 
               fun.ymax = function(x) mean(x) + sd(x), 
               geom = "ribbon", alpha=0.3, colour=NA) +
  stat_summary(fun.y = mean, geom = "line") + 
  coord_cartesian(xlim = c(0, .3)) + 
  facet_wrap(~ timeseries + statistic, scale="free_y") + 
  scale_color_discrete(labels=relabel, name="Penalty function") + 
  scale_fill_discrete(labels=relabel, name="Penalty function") + 
  xlab("Penalty size (as fraction of NPV0)")
```

@@

@@

```{r Figure_5, fig.cap="",  dependson="histogram-calc", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

@@

```{r Figure_6, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}

@@

@@

```{r Figure_7, fig.cap="", dev=c("pdf", "svg", "png"), cache=FALSE, dev.args=list(bg="transparent")}
who <- c("penalty_fn", "ignore_fraction", "mismatched_fraction", "reduction")
table2 <- arrange(mismatches_df[who], reduction) 
names(table2) = c("penalty.fn", "ignoring", "mismatched", "reduction")
table2_long <- melt(table2, id = c('penalty.fn', 'reduction'))
table2_long$reduction <- as.factor(table2_long$reduction)

@@
